---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
stopifnot(file.exists(weights_file))
bebel_threads <- as.integer(Sys.getenv("BEBELM_NUM_THREADS", "8"))
if (is.na(bebel_threads) || bebel_threads < 1L) bebel_threads <- 8L
```

# Rbebelm

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rbebelm)](https://sounkou-bioinfo.r-universe.dev/Rbebelm)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`Rbebelm` is a focused R interface to
[`maximecb/bebelm`](https://github.com/maximecb/bebelm): local CPU inference for
Liquid AI LFM2.5-8B-A1B GGUF weights, exposed through a small R/Rust API.

The package provides model loading, tokenizer access, pooled embeddings, bounded
generation, persistent agents, BebeLM tool-call parsing, R tool dispatch, stream
events, and Rust-thread async jobs. It does not ship weights. Set
`BEBELM_WEIGHTS_FILE` to a local GGUF path or use the path directly.

## Install

```r
install.packages(
  "Rbebelm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

Source installs require Cargo/rustc and GNU make. Native builds compile scalar
and SIMD backend libraries where supported; the runtime dispatcher selects the
best installed backend for the current CPU.

## Load a model

```{r load}
library(Rbebelm)

model <- bebel_model_load(weights_file, num_threads = bebel_threads)
model
rbebelm_backend_features()
```

The GGUF is memory-mapped read-only by upstream BebeLM. Multiple agents created
from one `BebelModel` share the same in-process `Arc<Model>`. Async jobs created
from that model also share those weights while keeping independent caches.
Separate processes that load the same GGUF can share physical pages through the
operating-system page cache.

Threading is set when the model is loaded. `num_threads` initializes the
process-global Rayon pool once; generation and embedding calls use that pool.
Async calls add one Rust worker per job and can execute concurrently, so total
CPU demand is controlled by the model-load thread count and by how many jobs the
caller launches.

## Tokenizer and embeddings

```{r tokenize-embed}
ids <- bebel_tokenize(model, "Bamako", add_bos = FALSE)
ids
bebel_detokenize(model, ids)

emb <- bebel_embed(model, c(
  mali = "Bamako is the capital of Mali",
  italy = "Rome is the capital of Italy",
  france = "Paris is the capital of France"
))
dim(emb)
round(emb[1:2, 1:6], 3)

token_emb <- bebel_token_embed(model, "short stature", add_bos = FALSE)
token_emb
data.frame(
  token_index = token_emb$token_index,
  token_id = token_emb$ids,
  token = token_emb$tokens
)
```

## Generation

```{r generate}
events <- character()
answer <- bebel_generate(
  model,
  "The capital of France is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0,
  on_event = function(event) events <<- c(events, event$type)
)

answer
unique(events)
```

`on_event` is explicit. Pass `NULL` for quiet runs, a function for all events,
or a named list accepted by `bebel_event_handler()`.

```{r chat}
chat <- bebel_chat(
  model,
  "Answer in five words: what does mmap help with?",
  greedy = TRUE,
  max_gen = 24,
  max_think = 0,
  on_event = NULL
)
chat
```

## Agents

Agents keep transcript tokens, decode caches, and sampling settings. The loaded
weights are shared.

```{r agents}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 16, max_think = 0)
bebel_append_user(agent, "Say exactly: Paris noted.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "Say exactly: second turn complete.")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

turn1
turn2
bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
substr(bebel_transcript(agent), 1, 160)
```

The direct methods expose the same state:

```{r direct-methods}
length(agent$history())
identical(agent$history(), bebel_history(agent))
agent$clear()[c("history_tokens", "processed_tokens", "kv_tokens")]
```

## Tools

Tool declarations are typed S7 objects. Schemas are rendered into the system
turn that BebeLM expects, and parser support is delegated to upstream BebeLM for
the bracketed Pythonic format.

```{r tools}
lookup_capital <- bebel_tool(
  "lookup_capital",
  function(args, context, call) {
    context$calls <- c(context$calls, paste(call$name, args$country))
    c(Mali = "Bamako", Italy = "Rome", France = "Paris")[[args$country]]
  },
  description = "Return a capital city.",
  schema = list(
    properties = list(country = list(type = "string")),
    required = list("country")
  )
)

lookup_capital
bebel_tool_schema_json(lookup_capital)

ctx <- new.env(parent = emptyenv())
ctx$calls <- character()
call <- bebel_parse_tool_call('[lookup_capital(country="Mali")]')
Rbebelm:::invoke_bebel_tool(lookup_capital, call, ctx)
ctx$calls
```

## Async jobs

Async jobs use an aio-style surface: submit work, monitor queued events on the
R thread, then collect the completed `Turn`. `bebel_async_wait()` is the
blocking monitor used by model-level sync helpers, so stream callbacks always
run on the R thread. Several jobs can share one loaded model; the weights stay
shared and execution can overlap. Agent async jobs run on a cloned agent
snapshot: the original agent is not mutated.

```{r async}
job_a <- bebel_generate_async(
  model,
  "The capital of Italy is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0
)

job_b_agent <- bebel_agent(model, greedy = TRUE, max_gen = 8, max_think = 0)
bebel_append(job_b_agent, "The capital of Mali is")
job_b <- bebel_agent_generate_async(job_b_agent)

event_types_a <- character()
async_a <- bebel_async_wait(
  job_a,
  on_event = function(event) event_types_a <<- c(event_types_a, event$type)
)
async_b <- bebel_async_collect(job_b, wait = TRUE)

unique(event_types_a)
async_a
async_b
bebel_agent_info(job_b_agent)[c("history_tokens", "processed_tokens")]
```

## Generation benchmark

The generation benchmark uses bounded async batches against one loaded model.
It records per-job timings, token counts, event counts, and aggregate throughput
for reproducible CPU comparisons.

```{r benchmark}
prompts <- c(
  "The capital of Mali is",
  "The capital of Italy is",
  "The capital of Japan is"
)

bench <- bebel_benchmark_generation(
  model,
  prompts,
  concurrency = min(length(prompts), 2L),
  repeats = 1L,
  greedy = TRUE,
  max_gen = 8,
  max_think = 0
)

bench
bench$aggregate
bench$jobs[, c("prompt", "generated_tokens", "wall_seconds", "decode_tps", "event_count")]
```
