---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
```

# Rbebelm

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rbebelm)](https://sounkou-bioinfo.r-universe.dev/Rbebelm)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`Rbebelm` is both a generic R agent framework and a concrete native local-model
backend.

- The framework layer provides backend-agnostic `S7`/`s7contract` interfaces for
  LLM providers, extensions, skills, prompt templates, loop events,
  frontend/TUI command catalogs, and Pi-inspired append-only JSONL session trees.
- The concrete backend wraps upstream
  [`maximecb/bebelm`](https://github.com/maximecb/bebelm), a pure-Rust CPU-only
  implementation of [Liquid AI LFM2.5-8B-A1B](https://www.liquid.ai/blog/lfm2-5-8b-a1b)
  inference. The R package uses
  [`savvy`](https://github.com/yutannihilation/savvy) for the R/Rust boundary and
  a runtime backend layout for portable SIMD dispatch.

The intended architecture is loop-first: the agent loop owns lifecycle, queues,
tools, extensions, events, and sessions; consoles, RPC servers, and future TUIs
consume that loop instead of owning agent business logic. BebeLM is the bundled
native provider, not a requirement of the framework contracts.

## Installation

Install from R-universe:

```r
install.packages(
  "Rbebelm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

R-universe can also publish Linux binaries for this universe. To prefer those
binaries on Linux, use the universe binary repository pattern:

```r
options(repos = c(
  Rbebelm = sprintf(
    "https://sounkou-bioinfo.r-universe.dev/bin/linux/noble-%s/%s/",
    R.version$arch,
    substr(getRversion(), 1, 3)
  ),
  CRAN = sprintf(
    "https://cran.r-universe.dev/bin/linux/noble-%s/%s/",
    R.version$arch,
    substr(getRversion(), 1, 3)
  )
))
install.packages("Rbebelm")
```

Source installs require Cargo/rustc and GNU make. On Linux, macOS, and Windows,
`Rbebelm` builds separate Rust backend libraries when possible: scalar, AVX2,
and AVX-512 on x86_64; scalar and NEON on arm64. The dispatcher selects the best
installed backend supported by the current CPU/runtime before loading model code.

The model weights are not bundled with the R package. Download the GGUF weights
from the upstream model source documented by
[`bebelm`](https://github.com/maximecb/bebelm), then pass the local path to
`bebel_model_load()`.

## Quick start

Set `BEBELM_WEIGHTS_FILE` to the local GGUF path, or replace `weights` with an
explicit file path. The README examples are evaluated when a local model file is
available during rendering.

```{r quick-start, eval=has_weights}
library(Rbebelm)

weights <- Sys.getenv("BEBELM_WEIGHTS_FILE", "LFM2.5-8B-A1B-Q4_K_M.gguf")
model <- bebel_model_load(weights, num_threads = 2)

# Agent-first API: one loaded model can back several conversations.
agent <- bebel_agent(model, greedy = TRUE, max_gen = 48, max_think = 16)

bebel_append_user(agent, "What is the capital of Mali? Answer briefly.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "What about Italy?")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

turn1
turn2
bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
```

A `BebelAgent` owns the token transcript and decode caches while sharing the
loaded model weights. Later turns only prefill newly appended tokens. The direct
methods `agent$history()`, `agent$transcript()`, and `agent$clear()` expose the
same operations as `bebel_history(agent)`, `bebel_transcript(agent)`, and
`bebel_clear(agent)`.

```{r agent-state, eval=has_weights}
length(agent$history())
substr(agent$transcript(), 1, 80)
identical(agent$history(), bebel_history(agent))

reset_info <- agent$clear()
reset_info[c("history_tokens", "processed_tokens", "kv_tokens")]
```

For an interactive terminal loop, call `bebel_live_console(agent)` or
`bebel_live_console(model)` in an R session.

```{r live-console}
#| eval: false

chat <- bebel_agent(model, max_gen = 256, max_think = 64)
bebel_live_console(chat)
#> ╔══════════════════════════════════════════════════════╗
#> ║  Entering BebeLM live console.                     ║
#> ║  Type /quit or /exit to return to R.               ║
#> ╚══════════════════════════════════════════════════════╝
#> >>> What is BebeLM?
#> <think>
#> The user asks for a brief explanation of BebeLM.
#> </think>
#> BebeLM is a small, CPU-focused local language model runtime written in Rust.
#>
#> >>> Why use the R agent API?
#> The R agent keeps the transcript and decode caches alive across turns while
#> sharing the loaded GGUF weights.
#>
#> >>> /quit
```

You can also create the console directly from a model:

```{r live-console-model}
#| eval: false
bebel_live_console(model, max_gen = 256, max_think = 64)
```

Convenience helpers are still available for simple calls. `bebel_chat()` wraps a
single ChatML user/assistant turn:

```{r chat-helper, eval=has_weights}
# on_event defaults to bebel_console_event(): thinking and text print live.
result <- bebel_chat(
  model,
  "In one concise sentence, what does runtime SIMD dispatch do?",
  greedy = TRUE,
  max_gen = 48,
  max_think = 16,
  on_event = bebel_console_event(),
  check_interrupt = TRUE
)

result
```

For plain text completion, use `bebel_generate()`: 

```{r raw-generate, eval=has_weights}
raw_result <- bebel_generate(
  model,
  "Runtime SIMD dispatch is useful because",
  greedy = TRUE,
  max_gen = 24,
  max_think = 16,
  on_event = bebel_console_event(),
  check_interrupt = TRUE
)
raw_result
```

Use `bebel_append_system()` for an upstream-rendered ChatML system turn. With no
tools, the low-level `bebel_append()` form below is equivalent apart from being
more explicit about the tokens. When `tools` are supplied, BebeLM renders its
`List of tools: [...]` system preamble.

```{r agent-system, eval=has_weights}
system_agent <- bebel_agent(model)
bebel_append_system(system_agent, "You are concise.")
bebel_transcript(system_agent)

raw_system_agent <- bebel_agent(model)
bebel_append(raw_system_agent, "<|im_start|>system\nYou are concise.<|im_end|>\n")
identical(bebel_transcript(system_agent), bebel_transcript(raw_system_agent))
```

Agents can also be driven at the lower level with raw text or token ids.

```{r agent-raw, eval=has_weights}
raw_agent <- bebel_agent(model, greedy = TRUE, max_gen = 16, max_think = 0)
bebel_append(raw_agent, "The capital of Mali is")
raw_turn <- bebel_agent_generate(raw_agent, on_event = NULL)

ids <- bebel_tokenize(model, " and its airport code is", add_bos = FALSE)
bebel_append_tokens(raw_agent, ids)
bebel_history(raw_agent)[1:8]
bebel_token_ids()[c("TOKEN_THINK", "TOKEN_TOOL_CALL_START", "TOKEN_TOOL_CALL_END")]
raw_turn$text
```

Tools can be orchestrated with an Agent run loop. The `context` object is private
to R tools and hooks; it is not sent to the model. A tool is dispatched only when
the model emits a BebeLM tool-call block, so prompts should describe the
available tools and the expected call format. The prompt below asks directly for
the tool-call form so the example exercises the dispatch path.

```{r tools, eval=has_weights}
ctx <- new.env(parent = emptyenv())
ctx$thread_id <- "thread-001"
ctx$log <- character()

tools <- list(
  lookup_capital = bebel_tool(
    "lookup_capital",
    function(args, context, call) {
      context$log <- c(context$log, paste("tool", call$name, args$country))
      c(Mali = "Bamako", Italy = "Rome")[[args$country]]
    },
    description = "Return a capital city for a country."
  )
)

hooks <- list(
  tool_request = function(call, context, ...) {
    context$log <- c(context$log, paste("request", call$name))
  },
  tool_result = function(call, result, context, ...) {
    context$log <- c(context$log, paste("result", call$name, result))
  }
)

tool_prompt <- paste(
  "Return exactly this tool call and no other text:",
  "lookup_capital({\"country\":\"Italy\"})"
)

agent <- bebel_agent(model, greedy = TRUE, max_gen = 64, max_think = 0)
bebel_append_user(agent, tool_prompt)
run <- bebel_agent_run(agent, tools = tools, context = ctx, hooks = hooks, max_steps = 2)
run
ctx$log
```

## Generic agent and frontend/TUI framework

The BebeLM bindings are one implementation of a more generic R agent/frontend framework.
The loop itself is backend-agnostic: it talks to objects that implement the
`BebelAgentBackend` S7/s7contract interface. BebeLM implements that contract today;
other local or remote providers can implement the same generics later.

The core backend contract is intentionally small:

- `bebel_backend_append_user(agent, message)`
- `bebel_backend_append_system(agent, message, tools = NULL)`
- `bebel_backend_append_tool_result(agent, content)`
- `bebel_backend_assistant_turn(agent, on_event, check_interrupt, stop_on_tool_call)`
- `bebel_backend_info(agent)`, `bebel_backend_transcript(agent)`, and `bebel_backend_clear(agent)`

`bebel_agent_loop()` owns policy, queues, event emission, tool dispatch, and
session persistence. A console, RPC server, or future Rust TUI should consume the
loop; it should not own agent logic. The queue semantics mirror Pi's vocabulary:
`bebel_loop_steer()`, `bebel_loop_follow_up()`, `steering_mode`, and
`follow_up_mode`.

```{r generic-session}
library(Rbebelm)

store <- bebel_session_create(
  cwd = tempdir(),
  session_dir = file.path(tempdir(), "rbebelm-readme-sessions"),
  name = "README demo"
)

user_id <- bebel_session_append_message(store, "user", "Hello from R")
bebel_session_append_message(
  store,
  "assistant",
  list(list(type = "text", text = "Hello.")),
  provider = "demo",
  model = "demo-model",
  stopReason = "stop"
)

bebel_session_leaf_id(store)
bebel_session_context(store)$messages
```

Session files are append-only JSONL trees inspired by Pi's session format. Each
entry has an `id` and `parentId`, so `/tree`, `/fork`, and `/clone` style UIs can
be built without rewriting history. By default, persisted sessions live under
`tools::R_user_dir("Rbebelm", "data")/sessions/<encoded-cwd>/`; set
`RBEBELM_SESSION_DIR` or pass `session_dir` to override that location.

Extensions are also generic capability bundles. An object implementing
`BebelAgentExtension` contributes a manifest plus optional tools, commands, hooks,
skill providers, prompt-template providers, and UI metadata. The loop registers
those capabilities; frontends only render and invoke the catalog.

```{r extension-provider-demo}
skills <- bebel_skill_provider(list(
  concise = "Prefer concise, direct answers."
))

prompts <- bebel_prompt_template_provider(list(
  system = "You are {{role}} working in {{place}}."
))

ext <- bebel_extension(
  "readme-demo",
  skill_providers = list(default = skills),
  prompt_template_providers = list(default = prompts),
  commands = list(info = bebel_loop_command("info", function(args, loop, context) {
    bebel_loop_state(loop)
  }))
)

bebel_extension_manifest(ext)
bebel_system_prompt(
  prompts,
  "system",
  data = list(role = "an R agent", place = "Bamako"),
  skill_provider = skills,
  skills = "concise"
)
```

See the "Generic agent and frontend framework" vignette for a fake backend
example and the full session-tree/forking API.

## Native FFF fuzzy file search

Rbebelm uses the FFF engine behind `fff-c` as its native fuzzy file search
primitive. A persistent `BebelFileFinder` indexes a project once; consoles, RPC
clients, default file tools, and future TUIs can reuse it for low-latency file
picking.

```{r fff-file-search}
search_root <- tempfile("rbebelm-fff-readme-")
dir.create(file.path(search_root, "src"), recursive = TRUE)
writeLines("demo", file.path(search_root, "src", "bamako_agent.R"))
writeLines("notes", file.path(search_root, "README.md"))

finder <- bebel_file_finder(search_root, watch = FALSE)
bebel_file_search(finder, "agent", limit = 5)
```

The native FFF dependency is not loaded in webR/wasm; the package still loads
there, and file-search creation reports that native FFF is unavailable. SIMD is
handled with the same care as the BebeLM backend: Rbebelm builds scalar and
optimized native dylibs separately, and FFF/`neo_frizbee` SIMD kernels are only
entered through runtime CPU-feature checks or the already-selected optimized
backend.

## R-native agent layer

`bebel_r_agent()` adds a small Corteza-inspired harness on top of the core
model bindings. One session object owns a BebeLM agent, a tool catalog, a private
context environment, and can be driven either from an R console or from a small
JSON-RPC SDK surface.

```{r r-agent-layer, eval=has_weights}
r_agent <- bebel_r_agent(
  model,
  allow_eval = FALSE,
  greedy = TRUE,
  max_gen = 96,
  max_think = 16
)
bebel_agent_tool_catalog(r_agent$tools)
```

Interactive console. The `/r` command is a direct R escape hatch into the same
environment used by the agent's R tools; for example `/r x <- mtcars` creates an
object that `r_objects()` can later see. Visible `/r` output is capped so large
objects do not flood the chat prompt; assign objects or use summaries such as
`/r str(x)` for inspection. For plots in an `Rscript` terminal, use `/rplot`,
e.g. `/rplot plot(mpg ~ cyl, mtcars)`, which saves a PNG under `rbebelm-plots/`.
The `r_eval` and `r_plot` tools are only advertised to the model when
`allow_eval = TRUE`.

```{r r-agent-console, eval=FALSE}
bebel_r_agent_console(r_agent)
```

For a one-call launcher from R, use `bebel_r_agent_start()`. It keeps the
loaded BebeLM model object local to the launcher while still sharing
`.GlobalEnv` with `/r` and the agent's R tools. The console prints a compact
stats line after each user turn.

```{r r-agent-console-oneliner, eval=FALSE}
bebel_r_agent_start(Sys.getenv("BEBELM_WEIGHTS_FILE", "LFM2.5-8B-A1B-Q4_K_M.gguf"))
```

The package also installs a small script in `inst/bin`:

```{r r-agent-bin, eval=FALSE}
agent_bin <- system.file("bin/rbebelm-agent", package = "Rbebelm")
system2(agent_bin, "--help")
```

From a shell, after installation:

```sh
"$(Rscript -e 'cat(system.file("bin/rbebelm-agent", package = "Rbebelm"))')" \
  --weights /path/to/LFM2.5-8B-A1B-Q4_K_M.gguf
```

Optional RPC server, using optional `nanonext`; JSON parsing and serialization use imported `yyjsonr`:

```{r r-agent-rpc, eval=FALSE}
server <- bebel_r_agent_rpc_server(r_agent, url = "http://127.0.0.1:8080")
server$start()
# ... handle requests ...
server$close()
```

The RPC endpoint accepts `POST /rpc` JSON-RPC calls such as `tools/list`,
`session/info`, `session/transcript`, `session/clear`, and `turn`.

The same event stream can be consumed programmatically. For example, collect
only answer-text deltas while suppressing console output:

```{r event-callback, eval=has_weights}
deltas <- character()
invisible(bebel_generate(
  model,
  "A text delta callback can",
  greedy = TRUE,
  max_gen = 12,
  max_think = 16,
  on_event = bebel_event_handler(
    text_delta = function(event) deltas <<- c(deltas, event$delta)
  )
))
paste0(deltas, collapse = "")
```

You can also pass a named list of event-specific handlers directly:

```{r event-list-callback, eval=has_weights}
counts <- c(text_delta = 0L, thinking_delta = 0L, done = 0L)
invisible(bebel_generate(
  model,
  "An event handler list can",
  greedy = TRUE,
  max_gen = 4,
  max_think = 16,
  on_event = list(
    text_delta = function(event) counts["text_delta"] <<- counts[["text_delta"]] + 1L,
    thinking_delta = function(event) counts["thinking_delta"] <<- counts[["thinking_delta"]] + 1L,
    done = function(event) counts["done"] <<- counts[["done"]] + 1L
  )
))
counts
```

## Interrupts and streaming

Generation checks `R_CheckUserInterrupt()` during prompt prefill and before every
decoded token, wrapped through savvy's unwind protection so Ctrl-C does not
longjmp through Rust frames. Streaming is event-based and uses a finite event
protocol; `bebel_event_types()` reports the event enum for this build. The
current enum is returned by `bebel_event_types()` and includes stream lifecycle,
thinking blocks, answer text blocks, tool-list blocks, tool-call blocks, and
`done`. Delta events contain `delta`, `id`, and `index`; control start/end events
include the delimiter token `id` and `marker`; end events contain accumulated
`content`. Console printing is just the default event handler. Use
`on_event = NULL` for silent batch generation.

## webR / wasm

`Rbebelm` builds the real Rust/savvy backend for webR as a static
`wasm_simd128` backend. The wasm build uses a patched local copy of upstream
BebeLM that avoids native-only `mmap` and Rayon imports on Emscripten: GGUF files
are read from the webR filesystem into memory and matmul runs serially. If you
mount or download a GGUF into the webR virtual filesystem, `bebel_model_load()`
will attempt to load it. Very large models can still exhaust browser/webR memory.

## Runtime backend dispatch

`Rbebelm` installs one small R shared library plus separate Rust backend
libraries. The R shared library owns registration and dispatch; model code lives
in the selected backend library. The dispatcher checks CPU/OS support before
loading SIMD backends, so a portable binary can avoid executing unsupported
instructions.

Backend selection happens once per R process. If you need to benchmark or debug
a specific backend, call `rbebelm_set_backend()` before the first native Rbebelm
call in a fresh `Rscript` process:

```{r request-auto, eval=FALSE}
rbebelm_set_backend("auto")
```

Inspect the current CPU/runtime and selected backend:

```{r backend-state, eval=has_weights}
rbebelm_cpuid_info()
rbebelm_backend_features()
rbebelm_backend_info()
```

## Development

Common development commands from the repository root. The `make vignettes`
target uses `rawvignette`; install it with
`remotes::install_github("matthewkling/rawvignette")` when editing
`vignettes-raw/`.

```sh
make rd           # regenerate savvy wrappers, dispatch init, NAMESPACE, and man/*.Rd
make rdm          # regenerate README.md from evaluated README.Rmd
make dev-install  # install the package locally from source
make test         # run tinytest tests
make check        # build and run R CMD check --no-manual
make vignettes    # precompile vignettes-raw/ into vignettes/ with rawvignette
make site         # build the pkgdown site
make clean        # remove generated build artifacts
```
