---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
embedding_weights_file <- Sys.getenv(
  "EMBEDDING_GEMMA_WEIGHTS_FILE",
  "/root/bebelm/embeddinggemma-300M-Q8_0.gguf"
)
stopifnot(file.exists(weights_file), file.exists(embedding_weights_file))
bebel_threads <- as.integer(Sys.getenv("BEBELM_NUM_THREADS", "8"))
if (is.na(bebel_threads) || bebel_threads < 1L) bebel_threads <- 8L
```

# Rbebelm

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rbebelm/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rbebelm)](https://sounkou-bioinfo.r-universe.dev/Rbebelm)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`Rbebelm` provides local pure-Rust CPU inference for two complementary GGUF
models: Liquid AI LFM2.5-8B-A1B through
[`maximecb/bebelm`](https://github.com/maximecb/bebelm), and Google's
retrieval-trained EmbeddingGemma-300M through a dedicated native encoder.

The package provides model loading, tokenizer access, semantic text embeddings,
causal contextual-state extraction, bounded generation, persistent agents,
BebeLM tool-call parsing, R tool dispatch, stream events, and Rust-thread async
jobs. It does not ship weights. Set `BEBELM_WEIGHTS_FILE` and
`EMBEDDING_GEMMA_WEIGHTS_FILE` to local GGUF paths or pass paths directly.
EmbeddingGemma weights are governed by the
[Gemma Terms of Use](https://ai.google.dev/gemma/terms).

## Install

```r
install.packages(
  "Rbebelm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

Source installs require Cargo/rustc and GNU make. Native builds compile scalar
and SIMD backend libraries where supported; the runtime dispatcher selects the
best installed backend for the current CPU.

## Load a model

```{r load}
library(Rbebelm)

model <- bebel_model_load(weights_file, num_threads = bebel_threads)
model
rbebelm_backend_features()
```

The GGUF is memory-mapped read-only by upstream BebeLM. Multiple agents created
from one `BebelModel` share the same in-process `Arc<Model>`. Async jobs created
from that model also share those weights while keeping independent caches.
Separate processes that load the same GGUF can share physical pages through the
operating-system page cache.

Threading is set when the model is loaded. `num_threads` initializes the
process-global Rayon pool once; generation and state-extraction calls use that pool.
Async calls add one Rust worker per job and can execute concurrently, so total
CPU demand is controlled by the model-load thread count and by how many jobs the
caller launches.

## Tokenizer and contextual states

```{r tokenize-states}
ids <- bebel_tokenize(model, "Bamako", add_bos = FALSE)
ids
bebel_detokenize(model, ids)

states <- bebel_pooled_states(model, c(
  mali = "Bamako is the capital of Mali",
  italy = "Rome is the capital of Italy",
  france = "Paris is the capital of France"
))
states

token_states <- bebel_token_states(model, "short stature")
token_states
data.frame(
  token_index = token_states$token_index,
  token_id = token_states$ids,
  token = token_states$tokens
)
```

These vectors are post-final-RMSNorm contextual states from a causal language
model. The default pooled representation uses SGPT-style position-weighted mean
pooling, which gives later states more weight because they have observed more of
the sequence. LFM2.5-8B-A1B was not trained as a semantic embedding or
late-interaction retrieval model, so `Rbebelm` labels these outputs as contextual
states and records `retrieval_trained = FALSE`. Do not use raw cosine or MaxSim
scores as production relevance scores without task-specific training and
evaluation.

## Retrieval-trained EmbeddingGemma

EmbeddingGemma is a separate stateless model handle. The Rust implementation
loads the `gemma-embedding` GGUF directly and runs its bidirectional encoder,
mean pooling, both learned dense projections, and L2 normalization without
linking to llama.cpp, PyTorch, ONNX Runtime, or the SentencePiece C++ library.

```{r embeddinggemma}
embedding_model <- embeddinggemma_model_load(
  embedding_weights_file,
  num_threads = bebel_threads
)

query_embedding <- embeddinggemma_embed_query(
  embedding_model,
  "capital of Mali"
)
document_embeddings <- embeddinggemma_embed_document(
  embedding_model,
  c(
    mali = "Bamako is the capital and largest city of Mali.",
    italy = "Rome is the capital city of Italy.",
    desert = "The Sahara is a desert in northern Africa."
  )
)

sort(drop(document_embeddings %*% as.numeric(query_embedding)), decreasing = TRUE)
```

The query and document helpers apply different prompts because that distinction
is part of the model's training contract. Character vectors use bounded packed
encoder batches while retaining independent positions and attention boundaries.
`embeddinggemma_embed()` requires an explicit task for other uses. Set
`dimensions` to 512, 256, or 128 for
Matryoshka truncation and re-normalization.

## Generation

```{r generate}
events <- character()
answer <- bebel_generate(
  model,
  "The capital of France is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0,
  on_event = function(event) events <<- c(events, event$type)
)

answer
unique(events)
```

`on_event` is explicit. Pass `NULL` for quiet runs, a function for all events,
or a named list accepted by `bebel_event_handler()`.

```{r chat}
chat <- bebel_chat(
  model,
  "Answer in five words: what does mmap help with?",
  greedy = TRUE,
  max_gen = 24,
  max_think = 0,
  on_event = NULL
)
chat
```

## Agents

Agents keep transcript tokens, decode caches, and sampling settings. The loaded
weights are shared.

```{r agents}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 16, max_think = 0)
bebel_append_user(agent, "Say exactly: Paris noted.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "Say exactly: second turn complete.")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

turn1
turn2
bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
substr(bebel_transcript(agent), 1, 160)
```

The direct methods expose the same state:

```{r direct-methods}
length(agent$history())
identical(agent$history(), bebel_history(agent))
agent$clear()[c("history_tokens", "processed_tokens", "kv_tokens")]
```

## Tools

Tool declarations are typed S7 objects. Schemas are rendered into the system
turn that BebeLM expects, and parser support is delegated to upstream BebeLM for
the bracketed Pythonic format.

```{r tools}
lookup_capital <- bebel_tool(
  "lookup_capital",
  function(args, context, call) {
    context$calls <- c(context$calls, paste(call$name, args$country))
    c(Mali = "Bamako", Italy = "Rome", France = "Paris")[[args$country]]
  },
  description = "Return a capital city.",
  schema = list(
    properties = list(country = list(type = "string")),
    required = list("country")
  )
)

lookup_capital
bebel_tool_schema_json(lookup_capital)

ctx <- new.env(parent = emptyenv())
ctx$calls <- character()
call <- bebel_parse_tool_call('[lookup_capital(country="Mali")]')
Rbebelm:::invoke_bebel_tool(lookup_capital, call, ctx)
ctx$calls
```

## Async jobs

Async jobs use an aio-style surface: submit work, monitor queued events on the
R thread, then collect the completed `Turn`. `bebel_async_wait()` is the
blocking monitor used by model-level sync helpers, so stream callbacks always
run on the R thread. Several jobs can share one loaded model; the weights stay
shared and execution can overlap. Agent async jobs run on a cloned agent
snapshot: the original agent is not mutated.

```{r async}
job_a <- bebel_generate_async(
  model,
  "The capital of Italy is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0
)

job_b_agent <- bebel_agent(model, greedy = TRUE, max_gen = 8, max_think = 0)
bebel_append(job_b_agent, "The capital of Mali is")
job_b <- bebel_agent_generate_async(job_b_agent)

event_types_a <- character()
async_a <- bebel_async_wait(
  job_a,
  on_event = function(event) event_types_a <<- c(event_types_a, event$type)
)
async_b <- bebel_async_collect(job_b, wait = TRUE)

unique(event_types_a)
async_a
async_b
bebel_agent_info(job_b_agent)[c("history_tokens", "processed_tokens")]
```

## Generation benchmark

The generation benchmark uses bounded async batches against one loaded model.
It records per-job timings, token counts, event counts, and aggregate throughput
for reproducible CPU comparisons.

```{r benchmark}
prompts <- c(
  "The capital of Mali is",
  "The capital of Italy is",
  "The capital of Japan is"
)

bench <- bebel_benchmark_generation(
  model,
  prompts,
  concurrency = min(length(prompts), 2L),
  repeats = 1L,
  greedy = TRUE,
  max_gen = 8,
  max_think = 0
)

bench
bench$aggregate
bench$jobs[, c("prompt", "generated_tokens", "wall_seconds", "decode_tps", "event_count")]
```
