---
title: "Generic agent and frontend framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generic agent and frontend framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`Rbebelm` is intentionally two things at once:

1. A generic, backend-agnostic R agent/frontend framework that can be implemented
   by BebeLM, another local model, or a remote provider.
2. A concrete native BebeLM backend for local GGUF inference.

The framework layer is deliberately small and interface-driven. The loop owns
agent lifecycle, queues, events, tool dispatch, extension registration, and JSONL
session persistence. Frontends such as a console, RPC server, or future Rust TUI
consume the same loop instead of reimplementing or owning agent logic.

## Backend contract

An LLM provider implements the `AgentBackend` interface from `s7contract` by
providing S7 methods for these generics:

- `agent_append_user(agent, message)`
- `agent_append_system(agent, message, tools = NULL)`
- `agent_append_tool_result(agent, content)`
- `agent_assistant_turn(agent, on_event, check_interrupt, stop_on_tool_call)`
- `agent_info(agent)`
- `agent_transcript(agent)`
- `agent_clear(agent)`

BebeLM implements this contract for `BebelAgent`. The following fake backend is
useful for tests and demonstrates that `bebel_agent_loop()` does not require a
BebeLM object.

```{r fake-backend}
library(Rbebelm)

FakeAgentBackendS3 <- S7::new_S3_class("fakeAgentBackendVignette")

S7::method(agent_append_user, FakeAgentBackendS3) <- function(agent, message) {
  agent$user <- c(agent$user, message)
  agent
}

S7::method(agent_append_system, FakeAgentBackendS3) <- function(agent, message, tools = NULL) {
  agent$system <- c(agent$system, message)
  agent
}

S7::method(agent_append_tool_result, FakeAgentBackendS3) <- function(agent, content) {
  agent$tool <- c(agent$tool, content)
  agent
}

S7::method(agent_assistant_turn, FakeAgentBackendS3) <- function(
  agent,
  on_event = NULL,
  check_interrupt = TRUE,
  stop_on_tool_call = FALSE
) {
  if (!is.null(on_event)) on_event(list(type = "text_delta", delta = "fake reply"))
  list(text = "fake reply", tokens = 2L, stop = "stop")
}

S7::method(agent_info, FakeAgentBackendS3) <- function(agent) {
  list(provider = "fake", model = "fake-model")
}

S7::method(agent_transcript, FakeAgentBackendS3) <- function(agent) {
  paste(c(agent$system, agent$user, agent$tool), collapse = "\n")
}

S7::method(agent_clear, FakeAgentBackendS3) <- function(agent) {
  agent$user <- character()
  agent$tool <- character()
  agent
}

backend <- structure(new.env(parent = emptyenv()), class = "fakeAgentBackendVignette")
backend$user <- character()
backend$system <- character()
backend$tool <- character()
```

## Loop and frontend ownership

`bebel_agent_loop()` is the UI-independent controller. It accepts any
`AgentBackend`, optional tools, hooks, extensions, and a persistence setting.
The queue vocabulary follows Pi:

- `bebel_loop_steer()` adds steering messages.
- `bebel_loop_follow_up()` adds follow-up messages.
- `bebel_loop_policy(steering_mode = ..., follow_up_mode = ...)` controls
  whether queued messages are drained one-at-a-time or all at once.

```{r loop-with-fake-backend}
store_dir <- file.path(tempdir(), "rbebelm-framework-sessions")
store <- agent_session_create(cwd = tempdir(), session_dir = store_dir, name = "fake backend")

loop <- bebel_agent_loop(backend, session = store)
run <- bebel_loop_run(loop, "Hello backend", max_steps = 1)

run$done
backend$user
bebel_loop_state(loop)[c("state", "turns", "session_file")]
```

The loop writes generic message entries to the session store. The backend keeps
its own transcript/cache state; the session file stores portable framework
history for replay, browsing, forking, sharing, and UI state.

```{r loop-session-context}
context <- agent_session_context(store)
vapply(context$messages, `[[`, character(1), "role")
readLines(agent_session_file(store), n = 3)
```

## JSONL session trees

Sessions are inspired by Pi's JSONL session format. The first line is a session
header; every other entry has an `id`, `parentId`, `timestamp`, and `type`.
Entries form a tree, not only a linear log. Moving the leaf to an earlier entry
and appending creates a new branch without deleting the old path.

Default persisted sessions live under:

```r
tools::R_user_dir("Rbebelm", "data")/sessions/<encoded-cwd>/
```

Set `RBEBELM_SESSION_DIR` or pass `session_dir` to override that location.

```{r session-tree}
s <- agent_session_create(cwd = tempdir(), session_dir = store_dir, name = "tree demo")
u1 <- agent_session_append_message(s, "user", "first question")
a1 <- agent_session_append_message(
  s,
  "assistant",
  list(list(type = "text", text = "first answer")),
  provider = "fake",
  model = "fake-model",
  stopReason = "stop"
)

agent_session_checkout(s, u1)
u2 <- agent_session_append_message(s, "user", "alternate branch")

vapply(agent_session_branch(s), `[[`, character(1), "type")
length(agent_session_tree(s)[[1]]$children)
```

Session entries include ordinary messages plus metadata and extension entries:

- `message`
- `custom` for extension state that does **not** enter model context
- `custom_message` for extension-injected context
- `label`
- `session_info`
- `model_change`
- `thinking_level_change`
- `compaction`
- `branch_summary`

```{r custom-session-entries}
agent_session_append_custom(s, "my-extension", list(counter = 1L))
agent_session_append_custom_message(s, "my-extension", "Hidden context", display = FALSE)
agent_session_append_label(s, u1, "checkpoint")

tail(vapply(agent_session_entries(s), `[[`, character(1), "type"), 3)
```

Forking copies all non-header entries into a new session file with a new header.
Cloning a branch copies only the active path from root to a selected leaf.

```{r fork-and-clone}
forked <- agent_session_fork(agent_session_file(s), cwd = tempdir(), session_dir = store_dir)
cloned <- agent_session_clone_branch(s, leaf_id = u2, session_dir = store_dir)

agent_session_header(forked)$parentSession
vapply(agent_session_entries(cloned), `[[`, character(1), "id")
```

## Extensions

An extension is a backend-agnostic capability bundle registered into the loop,
not into a particular terminal UI. It should implement the
`AgentExtension` interface:

- `extension_manifest(extension)`
- `extension_tools(extension)`
- `extension_commands(extension)`
- `extension_hooks(extension)`
- `extension_skill_providers(extension)`
- `extension_prompt_template_providers(extension)`

The helper `bebel_extension()` creates a simple extension object implementing
that interface. Extensions register into the loop; they do not own the loop or a
TUI.

```{r extension}
state_command <- bebel_loop_command(
  "state",
  function(args, loop, context) bebel_loop_state(loop),
  description = "Return loop state."
)

ext <- bebel_extension(
  "demo-extension",
  commands = list(state = state_command),
  hooks = list(event = function(event, loop, context, ...) {
    context$last_event_type <- event$type
  }),
  metadata = list(ui = "frontends may render this")
)

bebel_extension_manifest(ext)
```

When attached to a loop, contributed commands, tools, hooks, skill providers,
and prompt-template providers are available through loop state and catalogs.

```{r extension-loop}
loop2 <- bebel_agent_loop(backend, extensions = list(ext), session = FALSE)
bebel_loop_command_catalog(loop2)
bebel_loop_execute_command(loop2, "/state")
loop2$context$last_event_type
```

## Skills and prompt templates

Skill providers and prompt-template providers are separate interfaces so system
prompt loading is not tied to BebeLM. A provider can be in-memory, file-backed,
or package-backed.

```{r skill-template-providers}
skills <- agent_skill_provider(list(
  concise = "Prefer concise, direct answers.",
  r_safe = "Avoid side effects unless the user asks for them."
))

prompts <- agent_prompt_template_provider(list(
  system = "You are {{role}} working in {{place}}."
))

skill_list(skills)
prompt_template_list(prompts)

agent_system_prompt(
  prompts,
  "system",
  data = list(role = "an R coding agent", place = "Bamako"),
  skill_provider = skills,
  skills = c("concise", "r_safe")
)
```

`agent_append_system_prompt()` renders a template, appends selected skills, and
then calls `agent_append_system()`. This keeps system-prompt loading generic;
BebeLM-specific tool preamble details remain inside BebeLM's backend method.
