---
title: "Backends and webR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backends and webR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
```

`Rbebelm` builds native Rust backend libraries and dispatches to the best one
available for the current CPU. Backend selection happens before model code is
loaded.

```{r diagnostics}
rbebelm_cpuid_info()
rbebelm_backend_info()
rbebelm_backend_features()
```

The feature report includes the compiled SIMD flags and the storage policy used
for GGUF files.

```{r model-load}
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
stopifnot(file.exists(weights_file))
model <- bebel_model_load(weights_file, num_threads = 2)
model$info()
```

In native builds, upstream BebeLM maps the GGUF read-only. Multiple agents from
one model share weights through Rust `Arc<Model>`.

```{r shared-agents}
a <- bebel_agent(model, greedy = TRUE, max_gen = 8, max_think = 0)
b <- bebel_agent(model, greedy = TRUE, max_gen = 8, max_think = 0)

bebel_append(a, "The capital of Mali is")
bebel_append(b, "The capital of Italy is")

bebel_agent_generate(a, on_event = NULL)
bebel_agent_generate(b, on_event = NULL)
```

For webR, the same R API is the target. The browser runtime must provide a GGUF
in the webR filesystem and enough memory for the selected model.
