---
title: "Late-interaction retrieval with ColBERT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Late-interaction retrieval with ColBERT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
weights_file <- Sys.getenv("COLBERT_WEIGHTS_FILE", "/root/bebelm/LFM2.5-ColBERT-350M-Q4_K_M.gguf")
stopifnot(file.exists(weights_file))
model <- colbert_model_load(weights_file, num_threads = 2)
```

`LFM2.5-ColBERT-350M` is a retrieval-trained, non-causal encoder. Unlike a
dense embedding model, it preserves a matrix of vectors: one L2-normalized
128-dimensional row for each query or retained document token. Retrieval uses
ColBERT's late-interaction MaxSim operation:

\[
\operatorname{score}(q, d) = \sum_i \max_j q_i^\top d_j.
\]

## Query and document contracts

The native profile implements the release contract: `[Q] ` and `[D] ` prefixes,
a 32-position query (including learned PAD expansion positions), a 512-position
document, published document punctuation filtering, a learned dense projection,
and per-token L2 normalization. These are model semantics, not caller options.

```{r encode}
query <- colbert_encode_query(model, "capital of Mali")
document <- colbert_encode_document(
  model,
  "Bamako is the capital and largest city of Mali."
)

colbert_embeddings_info(query)
colbert_embeddings_info(document)
head(colbert_embedding_ids(document))
head(colbert_embedding_vectors(document))
colbert_maxsim(query, document)
```

## Exact candidate ranking

`colbert_rank()` is a small exact scorer for re-ranking candidate documents.
It deliberately does not claim to be an ANN or compressed late-interaction
index; a production corpus should be indexed by a system that preserves this
same model and preprocessing contract.

```{r rank}
colbert_rank(
  model,
  "capital of Mali",
  c(
    mali = "Bamako is the capital and largest city of Mali.",
    italy = "Rome is the capital city of Italy.",
    desert = "The Sahara is a desert in northern Africa."
  )
)
```
