---
title: "Contextual states and retrieval boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contextual states and retrieval boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
stopifnot(file.exists(weights_file))
model <- bebel_model_load(weights_file, num_threads = 2)
```

## What the package extracts

LFM2.5-8B-A1B is a causal text-generation model. `Rbebelm` can expose useful
intermediate features from it, but does not call those features semantic
embeddings.

For every token, `bebel_token_states()`:

1. runs the normal causal model, so a token sees itself and only its left context;
2. takes the residual state after the last transformer layer;
3. applies `token_embd_norm`, the model's final RMSNorm used immediately before
   the tied vocabulary projection; and
4. optionally L2-normalizes that token row.

```{r token-states}
tokens <- bebel_token_states(model, "Bamako is the capital of Mali")
tokens
head(data.frame(
  index = tokens$token_index,
  id = tokens$ids,
  token = tokens$tokens
))
```

`bebel_pooled_states()` pools the post-final-norm token states. Its default
`"weighted_mean"` mode uses weights `1, 2, ..., n`, then L2-normalizes the
result. Giving later causal states more weight follows the pooling baseline in
SGPT, where later states have observed more of the sequence.

```{r pooled-states}
states <- bebel_pooled_states(model, c(
  mali = "Bamako is the capital of Mali.",
  italy = "Rome is the capital of Italy."
))
states
attr(states, "state_info")
```

The extraction contract is explicit in `state_info`. In particular,
`retrieval_trained` is `FALSE`.

## What this does not establish

A mathematically valid cosine or MaxSim calculation is not evidence of a valid
retriever. The LFM weights have no contrastive retrieval objective, no trained
query/document encoding policy, and no late-interaction projection or
compression head. Consequently:

- cosine values from pooled states are uncalibrated;
- nearest-neighbor rankings can follow lexical or positional artifacts;
- raw token-state MaxSim can rank distractors above relevant passages; and
- adding natural-language instructions does not substitute for retrieval
  training and held-out evaluation.

The following computes the ColBERT-style *shape* of a late interaction for
inspection only. It is deliberately not exported as a retrieval API.

```{r maxsim-shape}
query <- bebel_token_states(model, "capital of Mali")$states
passage <- bebel_token_states(model, "Bamako is the capital of Mali")$states
similarity <- query %*% t(passage)
diagnostic_maxsim <- sum(apply(similarity, 1L, max))
diagnostic_maxsim
```

## A path to a real retriever

Turning these primitives into a supported retriever requires a separately
versioned, retrieval-trained artifact and evidence for its contract:

1. choose dense pooling or a late-interaction projection;
2. train on query-positive-negative examples with an appropriate contrastive
   or ranking loss;
3. define query and document formatting, special-token filtering, dimensions,
   and normalization as model metadata;
4. evaluate held-out retrieval datasets, hard negatives, languages, and target
   domains against established embedding baselines; and
5. only then expose an index/scoring API tied to that trained artifact.

Precedents include SGPT for weighted pooling of decoder states, LLM2Vec and
NV-Embed for adapting decoder-only models into encoders, and ColBERT for trained
token-level late interaction. Those systems support the direction, but they do
not validate raw LFM states by themselves.

## References

- Muennighoff (2022), [SGPT: GPT Sentence Embeddings for Semantic
  Search](https://arxiv.org/abs/2202.08904).
- BehnamGhader et al. (2024), [LLM2Vec](https://arxiv.org/abs/2404.05961).
- Lee et al. (2024), [NV-Embed](https://arxiv.org/abs/2405.17428).
- Khattab and Zaharia (2020), [ColBERT](https://arxiv.org/abs/2004.12832).
