---
title: "Retrieval embeddings with EmbeddingGemma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieval embeddings with EmbeddingGemma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
weights_file <- Sys.getenv(
  "EMBEDDING_GEMMA_WEIGHTS_FILE",
  "/root/bebelm/embeddinggemma-300M-Q8_0.gguf"
)
stopifnot(file.exists(weights_file))
model <- embeddinggemma_model_load(weights_file, num_threads = 2)
```

## A separate retrieval model

`BebelModel` and `EmbeddingGemmaModel` deliberately have different APIs.
LFM2.5 is a causal generator whose intermediate states are exposed only as
contextual features. EmbeddingGemma is a retrieval-trained, bidirectional text
encoder whose final vectors are intended for cosine similarity, retrieval,
clustering, and related representation tasks.

The package's EmbeddingGemma path is implemented in Rust. It reads the GGUF and
its tokenizer metadata directly, then runs:

1. BOS/content/EOS SentencePiece tokenization;
2. 24 bidirectional Gemma layers, using five symmetric-window layers followed
   by one global-attention layer;
3. final RMSNorm and attention-mask mean pooling;
4. the learned 768-to-3072 and 3072-to-768 projections; and
5. optional Matryoshka truncation followed by L2 normalization.

No llama.cpp, PyTorch, ONNX Runtime, or SentencePiece C++ library is linked into
the package. Oracle tests use the pinned `ggml-org` Q8_0 artifact at revision
`0f741b5a6585bd53aeb15cd1372c56f2a0f65e12` (SHA-256
`b5ce9d77a3fc4b3b39ccb5643c36777911cc4eb46a66962eadfa3f5f60490d63`).

## Query and document prompts

EmbeddingGemma was trained with asymmetric retrieval prompts. Use the dedicated
helpers rather than embedding unformatted query and document strings.

```{r retrieval}
query <- embeddinggemma_embed_query(model, "Which city is the capital of Mali?")
passages <- c(
  mali = "Bamako is the capital and largest city of Mali.",
  italy = "Rome is the capital city of Italy.",
  desert = "The Sahara is a desert in northern Africa."
)
documents <- embeddinggemma_embed_document(model, passages)

scores <- drop(documents %*% as.numeric(query))
sort(scores, decreasing = TRUE)
```

Character-vector inputs are packed into bounded encoder batches to reuse model
matrix reads. Position and attention boundaries are reset for every element, so
a batched result is numerically equivalent to encoding each string separately.

Document titles are optional. `NULL` applies the model's literal `title: none`
prompt; a title scalar is recycled over the input vector.

```{r titles}
with_titles <- embeddinggemma_embed_document(
  model,
  passages[c("mali", "italy")],
  title = c("Mali", "Italy")
)
with_titles
```

## Other tasks and Matryoshka dimensions

The generic function requires an explicit task so it cannot silently mix
retrieval-document, retrieval-query, or semantic-similarity prompts.

```{r semantic-matryoshka}
similarity_embeddings <- embeddinggemma_embed(
  model,
  c("A pleasant day", "Beautiful weather"),
  task = "semantic_similarity",
  dimensions = 256L
)

similarity_embeddings %*% t(similarity_embeddings)
attr(similarity_embeddings, "embedding_info")
```

Supported dimensions are 768, 512, 256, and 128. Lower-dimensional outputs are
the leading dimensions of the learned 768-vector and are re-normalized. Index
and query vectors must always use the same task contract, model artifact,
dimension, and normalization setting.

## Truncation and provenance

The model context is 2048 tokens including BOS and EOS. By default, longer
inputs are truncated while preserving those special tokens; set
`truncate = FALSE` to fail instead. Every result records per-input token counts
and truncation flags in `embedding_info`.

The package does not distribute model weights. EmbeddingGemma weights remain
subject to the [Gemma Terms of Use](https://ai.google.dev/gemma/terms) and
[Gemma Prohibited Use Policy](https://ai.google.dev/gemma/prohibited_use_policy).
