---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
colbert_weights_file <- Sys.getenv("COLBERT_WEIGHTS_FILE", "")
stopifnot(file.exists(weights_file))
colbert_available <- nzchar(colbert_weights_file) && file.exists(colbert_weights_file)
model <- bebel_model_load(weights_file, num_threads = 2)
```

`Rbebelm` loads local BebeLM generation and ColBERT late-interaction GGUF
profiles, then runs bounded CPU inference from R.

```{r diagnostics}
rbebelm_backend_info()
rbebelm_backend_features()
```

Tokenization is a direct BebeLM model operation.

```{r tokenize}
ids <- bebel_tokenize(model, "Bamako", add_bos = FALSE)
ids
bebel_detokenize(model, ids)
```

A real late-interaction retriever uses a distinct retrieval-trained profile,
with query/document token vectors and ColBERT MaxSim scoring.

```{r colbert, eval=colbert_available}
colbert <- colbert_model_load(colbert_weights_file, num_threads = 2)
colbert_rank(
  colbert,
  "capital of Mali",
  c(
    mali = "Bamako is the capital of Mali.",
    italy = "Rome is the capital of Italy."
  )
)
```

Set `COLBERT_WEIGHTS_FILE` to a local `LFM2.5-ColBERT-350M` GGUF when building
this vignette to execute the late-interaction example.

Generation returns text, token ids, stop reason, and timing statistics.

```{r generation}
out <- bebel_generate(
  model,
  "The capital of France is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0,
  on_event = NULL
)
out
```

A persistent agent keeps transcript and decode caches across turns while sharing
model weights with every other agent created from the same `BebelModel`.

```{r agent}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 12, max_think = 0)
bebel_append_user(agent, "Say exactly: Paris noted.")
bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "Say exactly: second turn complete.")
bebel_assistant_turn(agent, on_event = NULL)

bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
```
