---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rbebelm)
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE", "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf")
stopifnot(file.exists(weights_file))
model <- bebel_model_load(weights_file, num_threads = 2)
```

`Rbebelm` loads a local BebeLM GGUF, exposes tokenizer and contextual-state
primitives, and runs bounded CPU generation from R.

```{r diagnostics}
rbebelm_backend_info()
rbebelm_backend_features()
```

Tokenization and contextual-state extraction are direct model operations.

```{r contextual-states}
ids <- bebel_tokenize(model, "Bamako", add_bos = FALSE)
ids
bebel_detokenize(model, ids)

states <- bebel_pooled_states(model, c("Mali capital", "Italy capital"))
states
```

`bebel_pooled_states()` applies the model's final output norm before pooling;
`bebel_token_states()` returns those post-norm states token by token. These are
causal language-model features, not retrieval-trained semantic embeddings.
Their cosine or late-interaction behavior must be validated for each task.

Generation returns text, token ids, stop reason, and timing statistics.

```{r generation}
out <- bebel_generate(
  model,
  "The capital of France is",
  greedy = TRUE,
  max_gen = 8,
  max_think = 0,
  on_event = NULL
)
out
```

A persistent agent keeps transcript and decode caches across turns while sharing
model weights with every other agent created from the same `BebelModel`.

```{r agent}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 12, max_think = 0)
bebel_append_user(agent, "Say exactly: Paris noted.")
bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "Say exactly: second turn complete.")
bebel_assistant_turn(agent, on_event = NULL)

bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
```
