---
title: "Streaming generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Streaming generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Generation is event-based. The Rust decode loop emits a finite event protocol
reported by `bebel_event_types()`: stream start, BebeLM thinking block events,
answer text block events, tool-list/tool-call block events derived from special
token ids, and final done. Console streaming is implemented as an event handler,
`bebel_console_event()`, not as a separate model primitive.

```{r}
library(Rbebelm)
weights <- Sys.getenv("BEBELM_WEIGHTS_FILE")
if (nzchar(weights) && file.exists(weights)) {
  model <- bebel_model_load(weights, num_threads = 2)
  out <- bebel_generate(
    model,
    "The key benefit of runtime SIMD dispatch is",
    max_gen = 48,
    on_event = bebel_console_event(),
    check_interrupt = TRUE
  )
  out[c("stop", "generated_tokens", "decode_tps")]
} else {
  message("Set BEBELM_WEIGHTS_FILE to run the streaming example.")
}
```

Use `on_event = NULL` for silent batch-style generation, or provide a callback to
forward answer-text or thinking deltas to another system:

```{r}
text <- character()
thinking <- character()
# bebel_generate(model, "prompt", on_event = bebel_event_handler(
#   text_delta = function(event) text <<- c(text, event$delta),
#   thinking_delta = function(event) thinking <<- c(thinking, event$delta)
# ))
```

Adapters can forward the finite event stream to other interfaces. For example,
an SSE endpoint can serialize each event as it arrives, and chat integrations can
consume `text_delta`, `thinking_delta`, and `tool_call_delta` events.

Tool dispatch is driven by `tool_call_end` events. `bebel_agent_run()` does not
inspect ordinary answer text for tool calls; the model must emit a BebeLM
tool-call block, or the caller must provide a custom parser/adapter for the
model-specific format.
