---
title: "Agents and tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agents and tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
```

`Rbebelm` is organized around persistent agents. A `BebelAgent` keeps token
history, generation settings, and decode caches across turns. Tools are layered
on top of agents: the model emits a BebeLM tool-call block, R parses the call,
runs the matching R function, appends the tool result, and continues generation.

## Agent state

```{r, eval=has_weights}
library(Rbebelm)
model <- bebel_model_load(Sys.getenv("BEBELM_WEIGHTS_FILE"), num_threads = 2)
agent <- bebel_agent(model, greedy = TRUE, max_gen = 48, max_think = 16)

bebel_append_user(agent, "Say exactly: Paris noted.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "Say exactly: second turn complete.")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]

# Direct methods are available on the agent object.
length(agent$history())
substr(agent$transcript(), 1, 80)

# Helper functions provide the same operations.
length(bebel_history(agent))
substr(bebel_transcript(agent), 1, 80)

# Reset the conversation while keeping the loaded weights and generation settings.
agent$clear()[c("history_tokens", "processed_tokens", "kv_tokens")]
```

Use `bebel_append_system()` for a ChatML system-role instruction. Raw appends do
not add user framing, so the low-level `bebel_append()` form below is equivalent
apart from being more explicit about the tokens.

```{r, eval=has_weights}
system_agent <- bebel_agent(model)
bebel_append_system(system_agent, "You are concise.")
bebel_transcript(system_agent)

raw_system_agent <- bebel_agent(model)
bebel_append(raw_system_agent, "<|im_start|>system\nYou are concise.<|im_end|>\n")
identical(bebel_transcript(system_agent), bebel_transcript(raw_system_agent))

raw_agent <- bebel_agent(model, greedy = TRUE, max_gen = 16, max_think = 0)
bebel_append(raw_agent, "The capital of Mali is")
raw_turn <- bebel_agent_generate(raw_agent, on_event = NULL)
raw_turn[c("stop", "generated_tokens")]

ids <- bebel_tokenize(model, " and Italy is", add_bos = FALSE)
bebel_append_tokens(raw_agent, ids)
bebel_history(raw_agent)[1:8]
```

## Tool definitions

A tool is an R function plus metadata. The `context` environment is private to R
and is not appended to the model transcript.

```{r}
library(Rbebelm)
ctx <- new.env(parent = emptyenv())
ctx$thread_id <- "thread-001"
ctx$log <- character()

tools <- list(
  lookup_capital = bebel_tool(
    "lookup_capital",
    function(args, context, call) {
      context$log <- c(context$log, paste("tool", call$name, args$country))
      c(Mali = "Bamako", Italy = "Rome")[[args$country]]
    },
    description = "Return a capital city for a country."
  )
)

tools$lookup_capital
```

The default parser accepts JSON calls, simple function calls, and BebeLM's
bracketed tool-call form.

```{r}
bebel_parse_tool_call('{"name":"lookup_capital","arguments":{"country":"Italy"}}')
bebel_parse_tool_call('lookup_capital({"country":"Italy"})')
bebel_parse_tool_call('[lookup_capital(country="Italy")]')
```

## Run a tool loop

`bebel_agent_run()` dispatches tools only when generation emits a BebeLM
`tool_call_end` event. The prompt below asks directly for the tool-call form so
the example exercises the dispatch path.

```{r, eval=has_weights}
hooks <- list(
  tool_request = function(call, context, ...) {
    context$log <- c(context$log, paste("request", call$name))
  },
  tool_result = function(call, result, context, ...) {
    context$log <- c(context$log, paste("result", call$name, result))
  }
)

tool_prompt <- paste(
  "Return exactly this tool call and no other text:",
  "lookup_capital({\"country\":\"Italy\"})"
)

agent <- bebel_agent(model, greedy = TRUE, max_gen = 64, max_think = 0)
bebel_append_user(agent, tool_prompt)
run <- bebel_agent_run(agent, tools = tools, context = ctx, hooks = hooks, max_steps = 2)

length(run$tool_calls)
ctx$log
```

If a model uses a different tool-call format, pass a custom `parse_tool_call`
function to `bebel_agent_run()`.
