---
title: "Backends and webR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backends and webR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`Rbebelm` uses runtime backend dispatch so a portable R package can load a backend
that matches the current platform. The R shared library owns registration and
dispatch; model code lives in Rust backend libraries.

```{r}
library(Rbebelm)
rbebelm_cpuid_info()
rbebelm_backend_info()
```

## Backend selection

Backend loading happens once per R process. If you want to request a backend,
call `rbebelm_set_backend()` before the first call that loads backend symbols.

```{r}
rbebelm_set_backend("auto")
rbebelm_backend_features()
rbebelm_backend_info()
```

Supported backend names depend on the platform. Typical native builds include:

- `scalar`
- `avx2` and `avx512` on x86_64 when built
- `neon` on arm64 when built

If the requested backend is not installed or not supported by the current CPU,
the dispatcher reports an error before model code is loaded.

## webR

The webR build links a static `wasm_simd128` Rust backend. It uses a patched local
copy of upstream BebeLM for Emscripten:

- GGUF files are read from the webR filesystem into memory instead of using
  native `mmap`.
- Matmul runs serially because native Rayon threading is not used in webR.
- `bebel_model_load()` attempts to load a GGUF path from the webR virtual
  filesystem.

Very large GGUF files can exhaust browser or webR memory. Use smaller models or
browser/runtime settings appropriate for the target deployment.

The webR diagnostics report the static backend:

```{r, eval=FALSE}
rbebelm_backend_info()
rbebelm_backend_features()
```
