---
title: "Events and adapters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Events and adapters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
```

Generation in `Rbebelm` is event-based. The Rust decode loop emits a finite event
protocol reported by `bebel_event_types()`.

```{r}
library(Rbebelm)
bebel_event_types()
```

Events describe stream lifecycle, thinking blocks, answer text blocks,
tool-list blocks, tool-call blocks, and completion. Console output is just one
event handler: `bebel_console_event()`.

## Collect text deltas

Use `on_event = NULL` for silent generation, or provide a callback to consume
selected events.

```{r, eval=has_weights}
model <- bebel_model_load(Sys.getenv("BEBELM_WEIGHTS_FILE"), num_threads = 2)
text <- character()
thinking <- character()

turn <- bebel_generate(
  model,
  "A text delta callback can",
  greedy = TRUE,
  max_gen = 12,
  max_think = 16,
  on_event = bebel_event_handler(
    text_delta = function(event) text <<- c(text, event$delta),
    thinking_delta = function(event) thinking <<- c(thinking, event$delta)
  )
)

paste0(text, collapse = "")
turn[c("stop", "generated_tokens")]
```

## Handler lists

A named list can also be supplied directly. Unknown handler names are rejected so
stale event code fails early.

```{r, eval=has_weights}
counts <- c(text_delta = 0L, thinking_delta = 0L, done = 0L)

invisible(bebel_generate(
  model,
  "An event handler list can",
  greedy = TRUE,
  max_gen = 4,
  max_think = 16,
  on_event = list(
    text_delta = function(event) counts["text_delta"] <<- counts[["text_delta"]] + 1L,
    thinking_delta = function(event) counts["thinking_delta"] <<- counts[["thinking_delta"]] + 1L,
    done = function(event) counts["done"] <<- counts[["done"]] + 1L
  )
))

counts
```

## SSE example

An SSE endpoint can serialize each event as it arrives. The sketch below shows the
adapter shape; the exact response object depends on the web framework.

```{r, eval=FALSE}
sse_handler <- function(response) {
  bebel_event_handler(
    default = function(event) {
      payload <- jsonlite::toJSON(event, auto_unbox = TRUE, null = "null")
      response$write(paste0("event: ", event$type, "\n"))
      response$write(paste0("data: ", payload, "\n\n"))
      response$flush()
    }
  )
}

bebel_agent_generate(agent, on_event = sse_handler(response))
```

Tool adapters use the same event stream. `bebel_agent_run()` listens for
`tool_call_end` events, parses their accumulated content, invokes R tools, and
appends tool results to the transcript.
