---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
```

`Rbebelm` is both a generic R agent/frontend framework and a concrete native
local-model backend. Start from the entry point that matches your use case:

1. **Generic framework**: implement or consume the `BebelAgentBackend`,
   extension, skill, prompt-template, command-catalog, event, and JSONL
   session-tree interfaces without loading a model.
2. **Concrete BebeLM backend**: load GGUF weights and create a persistent
   `BebelAgent` that owns transcript and decode-cache state.
3. **Native file search**: create a persistent FFF file finder for agent tools,
   consoles, RPC clients, and the standalone terminal TUI.
4. **Terminal frontend**: use the ARF-inspired native `rbebelm-tui` binary, which
   talks to an R-owned loop over the stream/command endpoint instead of owning
   model or tool logic.

```{r}
library(Rbebelm)
rbebelm_backend_info()
```

## Framework-only start

You can use the framework contracts without GGUF weights. The loop consumes any
object that implements `BebelAgentBackend`; extensions add tools, commands,
hooks, skill providers, and prompt-template providers; session trees persist
portable JSONL history.

```{r}
store <- bebel_session_create(
  cwd = tempdir(),
  session_dir = file.path(tempdir(), "rbebelm-getting-started-sessions"),
  name = "getting started"
)

msg <- bebel_session_append_message(store, "user", "Hello from a framework-only session")
bebel_session_leaf_id(store)
bebel_session_context(store)$messages[[1]]$role
```

For the full backend contract, loop queue semantics, extension catalog, and
session-tree/forking API, see the "Generic agent and frontend framework"
vignette.

## Local model start

Model weights are not bundled with the package. Set `BEBELM_WEIGHTS_FILE` to a
local GGUF path before running the model examples, or pass an explicit file path
to `bebel_model_load()`.

```{r, eval=has_weights}
model <- bebel_model_load(Sys.getenv("BEBELM_WEIGHTS_FILE"), num_threads = 2)
rbebelm_backend_features()[c("backend", "target_arch", "target_os")]
```

```{r, eval=FALSE}
model <- bebel_model_load("LFM2.5-8B-A1B-Q4_K_M.gguf", num_threads = 2)
```

A BebeLM agent owns the transcript and decode caches while sharing loaded model
weights.

```{r, eval=has_weights}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 48, max_think = 16)

bebel_append_user(agent, "What is the capital of Mali? Answer briefly.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "What about Italy?")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

turn1$text
turn2$text
bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
```

Use `bebel_clear(agent)` to reset transcript and caches without reloading the
model.

## Convenience calls

For simple calls, `bebel_chat()` creates a single ChatML user/assistant turn.
For raw prompt completion, use `bebel_generate()`.

```{r, eval=has_weights}
chat <- bebel_chat(
  model,
  "In one concise sentence, what does runtime backend dispatch do?",
  greedy = TRUE,
  max_gen = 48,
  max_think = 16,
  on_event = NULL
)
chat$text
```

## Native FFF file search

The file-search layer is native-only and does not require model weights. It uses
vendored FFF behind `bebel_file_finder()` and `bebel_file_search()`.

```{r}
root <- tempfile("rbebelm-file-search-")
dir.create(file.path(root, "R"), recursive = TRUE)
writeLines("demo", file.path(root, "R", "agent_loop.R"))

finder <- bebel_file_finder(root, watch = FALSE)
bebel_file_search(finder, "agent", limit = 3)$path
```

In webR/wasm the package still loads, but creating a file finder reports that
native FFF search is unavailable.

## ARF-inspired TUI module

The terminal TUI source lives in `src/rust/src/bin/rbebelm-tui.rs` and is
compiled during source installation. It follows ARF's model: Rust owns terminal
rendering, configuration, and transport client behavior; R owns the agent loop,
tools, extensions, sessions, and model backend.

```sh
TUI="$(Rscript -e 'cat(system.file("bin/rbebelm-tui", package = "Rbebelm"))')"
"$TUI" config init
"$TUI" run --weights /path/to/LFM2.5-8B-A1B-Q4_K_M.gguf

# Or, like `pi`, use the bare command after configuring weights:
BEBELM_WEIGHTS_FILE=/path/to/LFM2.5-8B-A1B-Q4_K_M.gguf "$TUI"
```

The default `run` mode starts the R host, waits for readiness, attaches chat,
and stops the host when the UI exits. Slash commands are handled before model
turns: `/quit` exits, `/help` and `Tab` show completions, `/graphics` inspects or
sets plot handling, and `/rplot` draws with the configured R graphics device.
Model-side `r_eval` and `r_plot` are enabled by default for local TUI hosts; use
`--no-eval` at startup or `/no-eval` at runtime to remove them from the model
tool catalog. The lower-level split mode
still consumes `GET /stream` events plus `POST /command` messages, with
`POST /rpc` kept as compatibility/control API. There is no duplicate TUI-owned
agent loop and no core `/reload`; R registers extensions and frontends refresh
local palettes/widgets when `catalog_changed` events arrive.

## Token helpers

```{r, eval=has_weights}
ids <- bebel_tokenize(model, "The capital of Italy is", add_bos = TRUE)
ids
bebel_detokenize(model, ids)
bebel_token_ids()[c("TOKEN_THINK", "TOKEN_TOOL_CALL_START", "TOKEN_TOOL_CALL_END")]
```
