---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
```

`Rbebelm` provides `R` bindings to `BebeLM`, a Rust implementation of local
`CPU` inference for [Liquid AI LFM2.5-8B-A1B](https://www.liquid.ai/blog/lfm2-5-8b-a1b) GGUF weights. Model weights are not
bundled with the package. Set `BEBELM_WEIGHTS_FILE` to a local GGUF path before
running the model examples.

```{r}
library(Rbebelm)
rbebelm_backend_info()
```

## Load a model

```{r, eval=has_weights}
model <- bebel_model_load(Sys.getenv("BEBELM_WEIGHTS_FILE"), num_threads = 2)
rbebelm_backend_features()[c("backend", "target_arch", "target_os")]
```

If `BEBELM_WEIGHTS_FILE` is not set, use the same code with an explicit file
path:

```{r, eval=FALSE}
model <- bebel_model_load("LFM2.5-8B-A1B-Q4_K_M.gguf", num_threads = 2)
```

## Use an agent

The main interface is `BebelAgent`. An agent owns the transcript and decode
caches while sharing the loaded model weights.

```{r, eval=has_weights}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 48, max_think = 16)

bebel_append_user(agent, "What is the capital of Mali? Answer briefly.")
turn1 <- bebel_assistant_turn(agent, on_event = NULL)

bebel_append_user(agent, "What about Italy?")
turn2 <- bebel_assistant_turn(agent, on_event = NULL)

turn1$text
turn2$text
bebel_agent_info(agent)[c("history_tokens", "processed_tokens", "kv_tokens")]
```

Use `bebel_clear(agent)` to reset transcript and caches without reloading the
model.

## Convenience calls

For simple calls, `bebel_chat()` creates a single ChatML user/assistant turn.
For raw prompt completion, use `bebel_generate()`.

```{r, eval=has_weights}
chat <- bebel_chat(
  model,
  "In one concise sentence, what does runtime backend dispatch do?",
  greedy = TRUE,
  max_gen = 48,
  max_think = 16,
  on_event = NULL
)
chat$text
```

## Token helpers

```{r, eval=has_weights}
ids <- bebel_tokenize(model, "The capital of Italy is", add_bos = TRUE)
ids
bebel_detokenize(model, ids)
bebel_token_ids()[c("TOKEN_THINK", "TOKEN_TOOL_CALL_START", "TOKEN_TOOL_CALL_END")]
```
