---
title: "Model evaluation with vitals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model evaluation with vitals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
has_vitals <- requireNamespace("vitals", quietly = TRUE)
num_threads <- as.numeric(Sys.getenv("BEBELM_NUM_THREADS", "2"))
```

This vignette shows how to evaluate the local
[Liquid AI LFM2.5-8B-A1B](https://www.liquid.ai/blog/lfm2-5-8b-a1b) GGUF with
[`vitals`](https://github.com/tidyverse/vitals). The goal is to evaluate model
behavior on small, reproducible tasks, including tool use. It is not a
reproduction of Liquid AI's benchmark harnesses.

The source is in `vignettes-raw/`; the rendered file in `vignettes/` is
precompiled with `rawvignette` so `R CMD check`, pkgdown, and package installs do
not rerun the model.

```{r}
library(Rbebelm)
data.frame(
  model_file = if (has_weights) basename(weights_file) else NA_character_,
  has_weights = has_weights,
  has_vitals = has_vitals
)
```

If `vitals` is not installed, install it before regenerating this vignette:

```{r, eval=FALSE}
install.packages("vitals")
# or: pak::pak("tidyverse/vitals")
```

## Evaluation harness

`vitals::Task` expects solvers to return `result` and `solver_chat`. `Rbebelm` is
not yet an ellmer provider, so the helper below returns a minimal placeholder
object with class `"Chat"` for `vitals` bookkeeping while storing Rbebelm-specific
trace data in `solver_metadata`. The task is solved, scored, and measured; it is
not logged to the Inspect viewer in this vignette.

```{r harness, eval=has_weights && has_vitals}
library(vitals)

model <- bebel_model_load(weights_file, num_threads = num_threads)

rbebelm_vitals_chat <- function(model_name = "Rbebelm/LFM2.5-8B-A1B-GGUF") {
  structure(
    list(
      get_model = function() model_name,
      get_system_prompt = function() "",
      get_turns = function() list()
    ),
    class = "Chat"
  )
}

score_factor <- function(ok) {
  factor(ifelse(ok, "C", "I"), levels = c("I", "C"), ordered = TRUE)
}

vitals::vitals_log_dir_set(tempdir())

excerpt <- function(x, n = 120) {
  x <- gsub("\\s+", " ", x)
  ifelse(nchar(x) > n, paste0(substr(x, 1, n), "..."), x)
}

or_unknown <- function(x) {
  if (is.null(x) || length(x) == 0L || is.na(x)) "unknown" else x
}

run_task <- function(task, ...) {
  task$solve(...)
  task$score()
  task$measure()
  list(samples = task$get_samples(), metrics = task$metrics)
}
```

## Factual QA task

This task checks short deterministic factual answers. The scorer extracts an
`ANSWER:` line and compares it to the target.

```{r factual-task, eval=has_weights && has_vitals}
factual_data <- data.frame(
  input = c(
    "Return only 'ANSWER: Bamako'. What is the capital of Mali?",
    "Return only 'ANSWER: Rome'. What is the capital of Italy?",
    "Return only 'ANSWER: Tokyo'. What is the capital of Japan?"
  ),
  target = c("Bamako", "Rome", "Tokyo")
)

rbebelm_qa_solver <- function(inputs, model, max_gen = 48, ...) {
  result <- character(length(inputs))
  metadata <- vector("list", length(inputs))
  for (i in seq_along(inputs)) {
    agent <- bebel_agent(model, greedy = TRUE, max_gen = max_gen, max_think = 0)
    bebel_append_user(agent, inputs[[i]])
    turn <- bebel_assistant_turn(agent, on_event = NULL)
    result[[i]] <- turn$text
    metadata[[i]] <- list(
      chars = nchar(turn$text),
      history_tokens = bebel_agent_info(agent)$history_tokens
    )
  }
  list(
    result = result,
    solver_chat = lapply(inputs, function(...) rbebelm_vitals_chat()),
    solver_metadata = metadata
  )
}

answer_line_scorer <- function(samples) {
  extracted <- sub(".*ANSWER:\\s*([^\\n.]+).*", "\\1", samples$result, ignore.case = TRUE)
  matched <- grepl("ANSWER:", samples$result, ignore.case = TRUE) &
    trimws(tolower(extracted)) == trimws(tolower(samples$target))
  list(
    score = score_factor(matched),
    explanation = ifelse(matched, "matched ANSWER line", "missing or incorrect ANSWER line"),
    scorer_metadata = Map(function(answer, target) list(answer = answer, target = target), extracted, samples$target)
  )
}

qa_task <- vitals::Task$new(
  dataset = factual_data,
  solver = rbebelm_qa_solver,
  scorer = answer_line_scorer,
  name = "rbebelm-factual-qa"
)

qa_eval <- run_task(qa_task, model = model)
qa_eval$metrics
data.frame(
  sample = seq_len(nrow(qa_eval$samples)),
  target = qa_eval$samples$target,
  result = excerpt(qa_eval$samples$result),
  score = as.character(qa_eval$samples$score),
  explanation = qa_eval$samples$scorer_explanation,
  row.names = NULL
)
```

## Tool-use task

This task evaluates whether the model emits the requested tool call and whether
the agent loop reaches the expected final answer after tool execution. The tools
return values from a private R context, so the answer cannot be obtained by
calling an external service.

```{r tool-task, eval=has_weights && has_vitals}
tool_data <- data.frame(
  input = c(
    "Do not answer from memory. Emit exactly [lookup_capital(country=\"Mali\")]. After the tool result, answer exactly 'ANSWER: <tool result>'.",
    "Do not answer from memory. Emit exactly [lookup_currency(country=\"Mali\")]. After the tool result, answer exactly 'ANSWER: <tool result>'.",
    "Do not answer from memory. Emit exactly [lookup_capital(country=\"Italy\")]. After the tool result, answer exactly 'ANSWER: <tool result>'."
  ),
  target = c("Bamako", "XOF", "Rome"),
  expected_tool = c("lookup_capital", "lookup_currency", "lookup_capital")
)

rbebelm_tool_solver <- function(inputs, model, expected_tool, max_steps = 3, ...) {
  result <- character(length(inputs))
  metadata <- vector("list", length(inputs))

  for (i in seq_along(inputs)) {
    context <- new.env(parent = emptyenv())
    context$calls <- character()

    lookup_capital <- bebel_tool("lookup_capital", function(args, context) {
      country <- args$country
      context$calls <- c(context$calls, paste0("lookup_capital:", country))
      or_unknown(c(Mali = "Bamako", Italy = "Rome", Japan = "Tokyo")[[country]])
    })
    lookup_currency <- bebel_tool("lookup_currency", function(args, context) {
      country <- args$country
      context$calls <- c(context$calls, paste0("lookup_currency:", country))
      or_unknown(c(Mali = "XOF", Japan = "yen", Italy = "euro")[[country]])
    })

    agent <- bebel_agent(model, greedy = TRUE, max_gen = 128, max_think = 0)
    bebel_append_user(agent, inputs[[i]])
    run <- tryCatch(
      bebel_agent_run(
        agent,
        tools = list(lookup_capital, lookup_currency),
        context = context,
        max_steps = max_steps
      ),
      error = function(e) e
    )

    if (inherits(run, "error")) {
      result[[i]] <- bebel_transcript(agent)
      loop_error <- conditionMessage(run)
    } else {
      result[[i]] <- or_unknown(tail(run$turns, 1)[[1]]$text)
      if (identical(result[[i]], "unknown")) result[[i]] <- bebel_transcript(agent)
      loop_error <- NA_character_
    }
    metadata[[i]] <- list(
      calls = context$calls,
      call_count = length(context$calls),
      expected_tool = expected_tool[[i]],
      expected_tool_called = any(startsWith(context$calls, paste0(expected_tool[[i]], ":"))),
      loop_error = loop_error
    )
  }

  list(
    result = result,
    solver_chat = lapply(inputs, function(...) rbebelm_vitals_chat()),
    solver_metadata = metadata
  )
}

tool_scorer <- function(samples) {
  metadata <- samples$solver_metadata
  answer <- sub(".*ANSWER:\\s*([^\\n.]+).*", "\\1", samples$result, ignore.case = TRUE)
  answer_ok <- grepl("ANSWER:", samples$result, ignore.case = TRUE) &
    trimws(tolower(answer)) == trimws(tolower(samples$target))
  tool_ok <- vapply(metadata, function(x) isTRUE(x$expected_tool_called), logical(1))
  ok <- answer_ok & tool_ok
  list(
    score = score_factor(ok),
    explanation = ifelse(ok, "expected tool and answer observed", "missing expected tool call or final answer"),
    scorer_metadata = Map(
      function(answer, answer_ok, tool_ok) list(answer = answer, answer_ok = answer_ok, tool_ok = tool_ok),
      answer, answer_ok, tool_ok
    )
  )
}

tool_task <- vitals::Task$new(
  dataset = tool_data,
  solver = rbebelm_tool_solver,
  scorer = tool_scorer,
  name = "rbebelm-tool-use"
)

tool_eval <- run_task(tool_task, model = model, expected_tool = tool_data$expected_tool)
tool_eval$metrics
data.frame(
  sample = seq_len(nrow(tool_eval$samples)),
  target = tool_eval$samples$target,
  result = excerpt(tool_eval$samples$result),
  score = as.character(tool_eval$samples$score),
  answer_ok = vapply(tool_eval$samples$scorer_metadata, function(x) isTRUE(x$answer_ok), logical(1)),
  tool_ok = vapply(tool_eval$samples$scorer_metadata, function(x) isTRUE(x$tool_ok), logical(1)),
  calls = vapply(tool_eval$samples$solver_metadata, function(x) paste(x$calls, collapse = " | "), character(1)),
  explanation = tool_eval$samples$scorer_explanation,
  row.names = NULL
)
```

## Instruction-following task

This task checks simple constrained formatting. It is intentionally small; add
rows and epochs to make it more robust.

```{r instruction-task, eval=has_weights && has_vitals}
instruction_data <- data.frame(
  input = c(
    "Return exactly three comma-separated lowercase colors and nothing else.",
    "Return a JSON object with keys city and country for Bamako, and nothing else.",
    "Return exactly one line beginning with ANSWER: followed by the word ready."
  ),
  target = c("comma_colors", "json_city_country", "answer_ready")
)

rbebelm_instruction_solver <- function(inputs, model, ...) {
  result <- character(length(inputs))
  metadata <- vector("list", length(inputs))
  for (i in seq_along(inputs)) {
    agent <- bebel_agent(model, greedy = TRUE, max_gen = 80, max_think = 0)
    bebel_append_user(agent, inputs[[i]])
    turn <- bebel_assistant_turn(agent, on_event = NULL)
    result[[i]] <- turn$text
    metadata[[i]] <- list(chars = nchar(turn$text))
  }
  list(
    result = result,
    solver_chat = lapply(inputs, function(...) rbebelm_vitals_chat()),
    solver_metadata = metadata
  )
}

instruction_scorer <- function(samples) {
  ok <- logical(nrow(samples))
  for (i in seq_len(nrow(samples))) {
    x <- trimws(samples$result[[i]])
    ok[[i]] <- switch(
      samples$target[[i]],
      comma_colors = grepl("^[a-z]+,\\s*[a-z]+,\\s*[a-z]+$", x),
      json_city_country = grepl('^\\s*\\{.*"city"\\s*:', x) && grepl('"country"\\s*:', x),
      answer_ready = grepl("^ANSWER:\\s*ready\\s*$", x, ignore.case = TRUE),
      FALSE
    )
  }
  list(
    score = score_factor(ok),
    explanation = ifelse(ok, "format matched", "format did not match")
  )
}

instruction_task <- vitals::Task$new(
  dataset = instruction_data,
  solver = rbebelm_instruction_solver,
  scorer = instruction_scorer,
  name = "rbebelm-instruction-following"
)

instruction_eval <- run_task(instruction_task, model = model)
instruction_eval$metrics
data.frame(
  sample = seq_len(nrow(instruction_eval$samples)),
  target = instruction_eval$samples$target,
  result = excerpt(instruction_eval$samples$result),
  score = as.character(instruction_eval$samples$score),
  explanation = instruction_eval$samples$scorer_explanation,
  row.names = NULL
)
```

## Interpreting the results

These tasks evaluate observed model behavior under this GGUF, prompt set, backend,
and decoding configuration. They do not verify the architecture, training token
count, RL recipe, published benchmark numbers, or license terms. To make the
evaluation stronger, add more rows, set `epochs > 1`, vary decoding parameters,
and compare against other local or API-backed models with the same `vitals` task
definitions.
