---
title: "Model claims benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model claims benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
weights_file <- Sys.getenv("BEBELM_WEIGHTS_FILE")
has_weights <- nzchar(weights_file) && file.exists(weights_file)
num_threads <- as.numeric(Sys.getenv("BEBELM_NUM_THREADS", "2"))
context_probe_tokens <- as.numeric(Sys.getenv("BEBELM_CONTEXT_PROBE_TOKENS", "2048"))
`%maybe%` <- function(x, y) if (is.null(x) || length(x) == 0L || is.na(x)) y else x
```

This vignette records local probes related to public claims for
[Liquid AI LFM2.5-8B-A1B](https://www.liquid.ai/blog/lfm2-5-8b-a1b). It is a
runtime benchmark of the local GGUF and `Rbebelm` path, not a reproduction of
Liquid AI's training recipe or benchmark harnesses.

The source is in `vignettes-raw/`; the rendered file in `vignettes/` is
precompiled with `rawvignette` so `R CMD check`, pkgdown, and package installs do
not rerun the model.

```{r}
library(Rbebelm)
weights_file
has_weights
```

## Backend and model loading

```{r, eval=has_weights}
model <- bebel_model_load(weights_file, num_threads = num_threads)
rbebelm_backend_info()
rbebelm_backend_features()[c("backend", "target_arch", "target_os", "compiled_avx2", "compiled_dotprod")]
```

## Local generation throughput

This checks that the GGUF can run locally through the selected backend and records
throughput for a short deterministic completion.

```{r, eval=has_weights}
prompt <- "The capital of France is"
t0 <- proc.time()[["elapsed"]]
gen <- bebel_generate(
  model,
  prompt,
  greedy = TRUE,
  max_gen = 32,
  max_think = 0,
  on_event = NULL
)
t1 <- proc.time()[["elapsed"]]

data.frame(
  prompt = prompt,
  text = gen$text,
  generated_tokens = length(gen$tokens),
  elapsed_sec = round(t1 - t0, 3),
  tokens_per_sec = round(length(gen$tokens) / max(t1 - t0, .Machine$double.eps), 3),
  contains_paris = grepl("Paris", gen$text, ignore.case = TRUE)
)
```

## Tool dispatch

This probes the local agent loop: the model emits a BebeLM tool-call block, R
parses it, dispatches the R tool, appends the result, and continues the agent
transcript.

```{r, eval=has_weights}
agent <- bebel_agent(model, greedy = TRUE, max_gen = 96, max_think = 0)
bebel_append_user(agent, paste(
  "You may call exactly one tool. Do not answer from memory.",
  "To answer, emit exactly: [lookup_capital(country=\"Italy\")]"
))

context <- new.env(parent = emptyenv())
context$calls <- character()
hook_log <- character()

lookup_capital <- bebel_tool("lookup_capital", function(country, context) {
  context$calls <- c(context$calls, country)
  c(Italy = "Rome", France = "Paris")[[country]] %maybe% "unknown"
})

run <- bebel_agent_run(
  agent,
  tools = list(lookup_capital),
  context = context,
  max_steps = 2,
  hooks = list(
    tool_request = function(call, ...) hook_log <<- c(hook_log, paste("request", call$name)),
    tool_result = function(call, result, ...) hook_log <<- c(hook_log, paste("result", call$name, result))
  )
)

data.frame(
  tool_calls = length(run$tool_calls),
  context_calls = paste(context$calls, collapse = ","),
  hook_log = paste(hook_log, collapse = " | "),
  transcript_has_result = grepl("Rome", bebel_transcript(agent), fixed = TRUE)
)
```

## Two-step tool-chain probe

The public claims emphasize agentic tool use. This small probe asks for two tool
calls in sequence and reports what the local run actually did.

```{r, eval=has_weights}
chain_agent <- bebel_agent(model, greedy = TRUE, max_gen = 160, max_think = 0)
bebel_append_user(chain_agent, paste(
  "Use tools for each step.",
  "First emit exactly: [lookup_capital(country=\"France\")].",
  "After receiving that tool result, emit exactly: [lookup_country_info(city=\"Paris\")].",
  "Then answer briefly."
))

chain_context <- new.env(parent = emptyenv())
chain_context$calls <- character()

lookup_capital <- bebel_tool("lookup_capital", function(country, context) {
  context$calls <- c(context$calls, paste0("lookup_capital:", country))
  c(France = "Paris", Italy = "Rome")[[country]] %maybe% "unknown"
})

lookup_country_info <- bebel_tool("lookup_country_info", function(city, context) {
  context$calls <- c(context$calls, paste0("lookup_country_info:", city))
  c(Paris = "France", Rome = "Italy")[[city]] %maybe% "unknown"
})

chain <- bebel_agent_run(
  chain_agent,
  tools = list(lookup_capital, lookup_country_info),
  context = chain_context,
  max_steps = 4
)

data.frame(
  tool_calls = length(chain$tool_calls),
  calls = paste(chain_context$calls, collapse = " | "),
  chained_two_tools = length(chain_context$calls) >= 2L
)
```

## Non-Latin tokenization probe

This probes tokenizer behavior for a few non-Latin examples. It does not measure
language quality.

```{r, eval=has_weights}
texts <- c(
  japanese = "東京は日本の首都です。",
  arabic = "القاهرة مدينة كبيرة.",
  hindi = "दिल्ली भारत की राजधानी है।"
)

do.call(rbind, lapply(names(texts), function(name) {
  ids <- bebel_tokenize(model, texts[[name]], add_bos = FALSE)
  roundtrip <- bebel_detokenize(model, ids)
  data.frame(
    text = name,
    chars = nchar(texts[[name]], type = "chars"),
    tokens = length(ids),
    roundtrip_nonempty = nzchar(roundtrip)
  )
}))
```

## Context-path probe

The model announcement cites a 128K context window. This package-level probe does
not attempt a full 128K evaluation. It checks that a configurable long prompt can
pass through the local R/BebeLM path. Increase `BEBELM_CONTEXT_PROBE_TOKENS` for
heavier local experiments.

```{r, eval=has_weights}
seed <- "alpha beta gamma delta epsilon zeta eta theta "
long_prompt <- paste(rep(seed, ceiling(context_probe_tokens / 8)), collapse = "")
ids <- bebel_tokenize(model, long_prompt, add_bos = TRUE)
ids <- ids[seq_len(min(length(ids), context_probe_tokens))]
long_text <- bebel_detokenize(model, ids)

context_agent <- bebel_agent(
  model,
  greedy = TRUE,
  max_context = context_probe_tokens + 64,
  max_gen = 8,
  max_think = 0
)
bebel_append(context_agent, long_text)
long_out <- bebel_agent_generate(context_agent, on_event = NULL)

data.frame(
  requested_context_probe_tokens = context_probe_tokens,
  actual_prompt_tokens = length(ids),
  generated_tokens = length(long_out$tokens),
  context_path_ok = length(long_out$tokens) > 0L
)
```

## Optional `vitals` task

[`vitals`](https://github.com/tidyverse/vitals) is a general LLM evaluation
framework for R. When it is installed, the same local `Rbebelm` model can be
wrapped as a `vitals::Task` with a package-specific solver and scorer. This is
useful for turning the probes above into repeatable evaluation logs.

```{r, eval=has_weights && requireNamespace("vitals", quietly = TRUE)}
claim_dataset <- data.frame(
  input = c(
    "Answer with exactly: ANSWER: Paris. What is the capital of France?",
    "Answer with exactly: ANSWER: Rome. What is the capital of Italy?"
  ),
  target = c("Paris", "Rome")
)

rbebelm_solver <- function(inputs, model, ...) {
  result <- vapply(inputs, function(input) {
    agent <- bebel_agent(model, greedy = TRUE, max_gen = 32, max_think = 0)
    bebel_append_user(agent, input)
    bebel_assistant_turn(agent, on_event = NULL)$text
  }, character(1))
  list(
    result = result,
    solver_chat = vector("list", length(inputs)),
    solver_metadata = lapply(result, function(text) list(chars = nchar(text)))
  )
}

answer_scorer <- function(samples) {
  extracted <- sub(".*ANSWER:\\s*([^\\n.]+).*", "\\1", samples$result, ignore.case = TRUE)
  ok <- trimws(tolower(extracted)) == trimws(tolower(samples$target))
  list(
    score = factor(ifelse(ok, "C", "I"), levels = c("I", "C"), ordered = TRUE),
    scorer_metadata = Map(function(answer, target) list(answer = answer, target = target), extracted, samples$target)
  )
}

task <- vitals::Task$new(
  dataset = claim_dataset,
  solver = rbebelm_solver,
  scorer = answer_scorer,
  name = "rbebelm-capitals"
)

task$eval(model = model, view = FALSE)
task$get_samples()[c("input", "target", "result", "score")]
```

If `vitals` is not installed, install it with:

```{r, eval=FALSE}
install.packages("vitals")
# or: pak::pak("tidyverse/vitals")
```

## What this does and does not verify

```{r}
data.frame(
  claim_area = c(
    "local inference",
    "short-generation speed",
    "tool dispatch",
    "simple tool chaining",
    "non-Latin tokenizer path",
    "long-context plumbing",
    "8B/1.5B-active architecture",
    "38T training tokens and RL recipe",
    "published benchmark scores",
    "comparison to larger models",
    "license terms"
  ),
  probed_here = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
)
```
