---
title: "R-native agent layer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R-native agent layer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_weights <- nzchar(Sys.getenv("BEBELM_WEIGHTS_FILE")) && file.exists(Sys.getenv("BEBELM_WEIGHTS_FILE"))
```

RbebelM's low-level API exposes model loading, ChatML turns, events, and tool
orchestration. The R-native agent layer builds a small harness on top of those
pieces: one session object can be driven from a console loop or from a JSON-RPC
server.

The layer is intentionally separate from the core model bindings. The hot path
is still BebeLM/Rust inference; the R layer owns session policy, tool metadata,
console interaction, and RPC.

## Create a session

```{r, eval=has_weights}
library(Rbebelm)

model <- bebel_model_load(Sys.getenv("BEBELM_WEIGHTS_FILE"), num_threads = 2)
agent <- bebel_r_agent(
  model,
  allow_eval = FALSE,
  greedy = TRUE,
  max_gen = 128,
  max_think = 16
)
agent
```

The default tools are deliberately small:

```{r}
library(Rbebelm)
tools <- bebel_default_r_tools(allow_eval = FALSE)
bebel_agent_tool_catalog(tools)
```

They include R object/documentation inspection and read-only file tools. The
`r_eval` tool is only included when the agent is created with
`allow_eval = TRUE`.

## Run a turn

```{r, eval=has_weights}
turn <- bebel_r_agent_turn(
  agent,
  "Reply exactly: R agent ready.",
  max_steps = 1,
  on_event = NULL
)
trimws(sub("(?s)^.*</think>\\s*", "", turn$text, perl = TRUE))
```

For interactive use, start the console loop:

```{r, eval=FALSE}
bebel_r_agent_console(agent)
```

Or use the one-call launcher, which loads the model and then enters the console:

```{r, eval=FALSE}
bebel_r_agent_start(Sys.getenv("BEBELM_WEIGHTS_FILE", "LFM2.5-8B-A1B-Q4_K_M.gguf"))
```

The package also installs a small helper script in `inst/bin`:

```{r, eval=FALSE}
agent_bin <- system.file("bin/rbebelm-agent", package = "Rbebelm")
system2(agent_bin, "--help")
```

Console commands include `/tools`, `/r`, `/transcript`, `/clear`, and `/quit`.
The `/r` command evaluates R directly in the same environment used by the
agent's R tools. For example, `/r x <- mtcars` creates an object that
`r_objects()` can later see. Visible `/r` output is capped so large objects do
not flood the chat prompt; assign objects or use summaries such as `/r str(x)`
for inspection. The `r_eval` tool is only advertised to the model when the
session is created with `allow_eval = TRUE`. The console prints a compact
token/timing stats line after each user turn.

## JSON-RPC SDK surface

The same session object can be served over a small JSON-RPC API. This is not an
OpenAI-compatible API; it is an SDK surface for controlling the R-native agent.
`nanonext` and `jsonlite` are optional dependencies used only by this server.

```{r, eval=FALSE}
server <- bebel_r_agent_rpc_server(agent, url = "http://127.0.0.1:8080")
server$start()
# ... handle requests ...
server$close()
```

Supported methods:

- `session/info`
- `tools/list`
- `session/transcript`
- `session/clear`
- `turn`

Example request body for `POST /rpc`:

```json
{"jsonrpc":"2.0","id":1,"method":"turn","params":{"prompt":"List objects in the R session."}}
```

The console and RPC layers share the same session, transcript, tool catalog, and
BebeLM agent state.
